---
title: "Cyclistic bike-share analysis case study"
author: "Tram Anh Nguyen"
date: "2024-03-10"
output:
  pdf_document: default
  html_document: default
---
# DATA PROCESSING
## Load libraries
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
```
## Set my working directory and load 12 .csv files from Febuary 2023 to January 2024 
```{r}
setwd('/Users/9360/Desktop/Data Analytics/Google_Data_Analytics/Capstone_Project/Data')
 Feb2023_df <- read.csv("202302-divvy-tripdata.csv")
 Mar2023_df <- read.csv("202303-divvy-tripdata.csv")
 Apr2023_df <- read.csv("202304-divvy-tripdata.csv")
 May2023_df <- read.csv("202305-divvy-tripdata.csv")
 Jun2023_df <- read.csv("202306-divvy-tripdata.csv")
 Jul2023_df <- read.csv("202307-divvy-tripdata.csv")
 Aug2023_df <- read.csv("202308-divvy-tripdata.csv")
 Sep2023_df <- read.csv("202309-divvy-tripdata.csv")
 Oct2023_df <- read.csv("202310-divvy-tripdata.csv")
 Nov2023_df <- read.csv("202311-divvy-tripdata.csv")
 Dec2023_df <- read.csv("202312-divvy-tripdata.csv")
 Jan2024_df <- read.csv("202401-divvy-tripdata.csv")
```
## Merge all uploaded files into a new data frame 
```{r}
merged_df <- rbind(Feb2023_df, Mar2023_df, Apr2023_df, May2023_df, Jun2023_df, Jul2023_df, Aug2023_df, Sep2023_df, Oct2023_df, Nov2023_df, Dec2023_df, Jan2024_df)
```
## Remove 12 csv.files to keep the environment organized
```{r}
rm(Feb2023_df, Mar2023_df, Apr2023_df, May2023_df, Jun2023_df, Jul2023_df, Aug2023_df, Sep2023_df, Oct2023_df, Nov2023_df, Dec2023_df, Jan2024_df)
```
## Save the data frame as `merged_data` and get an overview of it 
```{r}
merged_data <- merged_df
glimpse(merged_data)
```
There are *13 variables* and *5,674,449 observations* in total. 

# DATA CLEANING AND MANIPULATION
## Data Cleaning 
### Remove duplicate rows 
```{r}
merged_data <- distinct(merged_data)
```
### Remove rows with NA values 
```{r}
merged_data <- na.omit(merged_data) 
```
### Remove rows with white space or empty string in any columns
```{r}
merged_data <- merged_data[rowSums(sapply(merged_data, function(x) x == "")) == 0, ]
```
### Remove unnecessary columns
```{r}
merged_data <- subset(merged_data, select = -c(ride_id, start_station_id, end_station_id, start_lat, start_lng, end_lat, end_lng))
```
### Save the data frame as `cleaned_data` and get an overview of it 
```{r}
cleaned_data <- merged_data
View(cleaned_data)
```

## Data Manipulation 
### Create new columns `Ride_length` that calculates the length of each ride by subtracting started_at column from ended_at column and converts it into minutes.
```{r}
# create the column
cleaned_data$ride_length <- difftime(cleaned_data$ended_at, cleaned_data$started_at, units = 'min')
cleaned_data$ride_length <- as.numeric(cleaned_data$ride_length) 

# check the first 5 rows
head(cleaned_data, 5)
```
### Create new column `Day_of_week` that extracts the day of the week from started_at column
```{r}
# create the column
cleaned_data$day_of_week <- wday(cleaned_data$started_at, label = TRUE, abbr = FALSE)

# check the first 5 rows
head(cleaned_data, 5)
```
### Create new column `Month` that extracts the month from started_at column
```{r}
# create the column
cleaned_data$month <- format(as.Date(cleaned_data$started_at), "%m")

# check the first 5 rows
head(cleaned_data, 5)
```
### Create new column `Day` that extracts the day from started_at column.
```{r}
# create the column
cleaned_data$day <- format(as.Date(cleaned_data$started_at), "%d") 

# check the first 5 rows
head(cleaned_data, 5)
```
### Create new column `Hour` that extracts the hour from started_at column
```{r}
# create the column
cleaned_data$hour <- format(as.POSIXct(cleaned_data$started_at), format = "%H")

# check the first 5 rows
head(cleaned_data, 5)
```
### Round_trip: identify whether the route is one-way or round-trip
```{r}
# create the column
cleaned_data$round_trip <- ifelse(cleaned_data$start_station_name == cleaned_data$end_station_name, 'Yes', 'No')

# check the first 5 rows
head(cleaned_data, 5)
```
## View the data frame 
```{r}
View(cleaned_data)
```
## Remove any rows which ride_length is negative
```{r}
cleaned_data <- cleaned_data %>%
  filter(ride_length >= 0)  
```
## Save the new data frame as `final_data`
```{r}
final_data <- cleaned_data
```

# DATA ANALYZATION
## Total Rides & Average Ride Length
```{r}
final_data %>%
 summarize(total_rides = nrow(final_data),average_ride_length = round(mean(ride_length), digits = 2))
```
## Total Rides and Average Ride Length by User Type 
```{r}
final_data %>%
  group_by(member_casual) %>%
  summarise(total_rides = length(member_casual),
            percentage = round((length(member_casual) / nrow(final_data)) * 100, digits = 2),
            average_ride_length = round(mean(ride_length), digits = 2))

```
## Total Rides and Average Ride Length by Rideable Type and User Type. 
```{r}
final_data %>%
  group_by(member_casual, rideable_type) %>%
  summarise(total_rides = length(member_casual),
            percentage = round((length(rideable_type) / nrow(final_data)) * 100, digits = 2),
            average_ride_length = round(mean(ride_length), digits = 2))
```
# DATA VISUALIZATION
## Total Rides by Day of the Week and User Type 
```{r}
final_data %>%
  group_by(member_casual, day_of_week) %>% 
  summarize(number_of_ride = n()) %>% 

ggplot(aes(factor(day_of_week,levels = c("Sunday", "Saturday", "Friday", "Thursday", "Wednesday", "Tuesday", "Monday")), number_of_ride, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.7) +
  facet_wrap(~ member_casual) + 
  geom_text(aes(label = format(number_of_ride, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 2.0) + 
  
  labs(x = "Day of week", y = "Number of Rides", 
       title = "Cyclistic - Total Rides (Feb 2023 - Jan 2024)", 
       subtitle =  "By Day of Week and User Type") +
  scale_fill_manual(values = c("#de6e56","#0b3c5d")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0, 550000)) +
  theme_classic(base_size = 10) +
  coord_flip()

ggsave("Total_Rides_by_Day_of_the_Week_and_User_Type.png", plot = last_plot(), dpi = 300)
```
## Average Ride Length by Day of the Week and User Type 
```{r}
final_data %>%
  group_by(member_casual, day_of_week) %>% 
  summarize(average_ride_length = round(mean(ride_length), digits = 2))%>% 
  
ggplot(aes(factor(day_of_week,levels = c("Sunday","Saturday", "Friday","Thursday","Wednesday","Tuesday","Monday")),average_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.7) +
  facet_wrap(~ member_casual) + 
  geom_text(aes(label = format(average_ride_length, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 2.0) + 
  
  labs(x = "Day of week", y = "Average Ride Length (min)", 
       title = "Cyclistic - Average Ride Length (Feb 2023 - Jan 2024)", 
       subtitle =  "By Day of Week and User Type") +
  scale_fill_manual(values = c("#de6e56","#0b3c5d")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0,30)) +
  theme_classic(base_size = 10) + 
  coord_flip()

ggsave("Average_Ride_Length_by_Day_of_the_Week_and_User_Type.png", plot = last_plot(), dpi = 300)
```
## Total Rides by Month and User Type
```{r}
final_data %>%
  group_by(member_casual, month) %>% 
  summarize(number_of_ride = n()) %>% 

ggplot(aes(month, number_of_ride, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.7) +
  facet_wrap(~ member_casual) + 
  geom_text(aes(label = format(number_of_ride, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 2.0) + 
  
  labs(x = "Month", y = "Number of Rides", 
       title = "Cyclistic - Total Rides (Feb 2023 - Jan 2024)", 
       subtitle =  "By Month and User Type") +
  scale_fill_manual(values = c("#de6e56","#0b3c5d")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0, 550000)) +
  theme_classic(base_size = 10) +
  coord_flip()

ggsave("Total_Rides_by_Month_and_User_Type.png", plot = last_plot(), dpi = 300)
```

## Average Ride Length by Month and User Type
```{r}
final_data %>%
  group_by(member_casual, month) %>% 
  summarize(average_ride_length = round(mean(ride_length), digits = 2))%>% 
  
ggplot(aes(month,average_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.7) +
  facet_wrap(~ member_casual) + 
  geom_text(aes(label = format(average_ride_length, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 2.0) + 
  
  labs(x = "Month", y = "Average Ride Length (min)", 
       title = "Cyclistic - Average Ride Length (Feb 2023 - Jan 2024)", 
       subtitle =  "By Month and User Type") +
  scale_fill_manual(values = c("#de6e56","#0b3c5d")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0,30)) +
  theme_classic(base_size = 10) + 
  coord_flip()

ggsave("Average_Ride_Length_by_Month_and_User_Type.png", plot = last_plot(), dpi = 300)
```
## Total Rides by User Type per Hour
```{r}
final_data %>%
  group_by(member_casual, hour) %>% 
  summarize(number_of_ride = n()) %>% 

ggplot(aes(hour, number_of_ride, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.7) +
  facet_wrap(~ member_casual) + 
  geom_text(aes(label = format(number_of_ride, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 2.0) + 
  
  labs(x = "Hour", y = "Number of Rides", 
       title = "Cyclistic - Total Rides (Feb 2023 - Jan 2024)", 
       subtitle =  "By Hour and User Type") +
  scale_fill_manual(values = c("#de6e56","#0b3c5d")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0, 350000)) +
  theme_classic(base_size = 10) +
  coord_flip()

ggsave("Total_Rides_by_Hour_and_User_Type.png", plot = last_plot(), dpi = 300)
```
## Average Ride Length by User Type per Hour
```{r}
final_data %>%
  group_by(member_casual, hour) %>% 
  summarize(average_ride_length = round(mean(ride_length), digits = 2))%>% 
  
ggplot(aes(hour,average_ride_length, fill = member_casual)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.7) +
  facet_wrap(~ member_casual) + 
  geom_text(aes(label = format(average_ride_length, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 2.0) + 
  
  labs(x = "Hour", y = "Average Ride Length (min)", 
       title = "Cyclistic - Average Ride Length (Feb 2023 - Jan 2024)", 
       subtitle =  "By Hour and User Type") +
  scale_fill_manual(values = c("#de6e56","#0b3c5d")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0,30)) +
  theme_classic(base_size = 10) + 
  coord_flip()

ggsave("Average_Ride_Length_by_Hour_and_User_Type.png", plot = last_plot(), dpi = 300)
```
## Total Rides & Average Ride Length of Round-trip and One-way Route by User Type
```{r}
 final_data %>%
  group_by(member_casual, round_trip) %>% 
  summarise(total_rides = length(round_trip),
            average_ride_length = round(mean(ride_length), digits = 2))
```
## Top 10 Most Frequent Routes for Member User
```{r}
final_data <- final_data %>%
  mutate(route = paste(start_station_name, "->", end_station_name))

# group by the new route column, filter for list of top 10 routes 
top_routes_member <- final_data %>%
  filter(member_casual == "member") %>%
  group_by(route, member_casual, round_trip) %>%
  summarize(number_of_ride = n()) %>%
  ungroup() %>%
  arrange(desc(number_of_ride)) %>% 
  slice_max(order_by = number_of_ride, n = 10)

# plot the top 10 routes
ggplot(top_routes_member, aes(x = reorder(route, number_of_ride), y = number_of_ride, fill = round_trip)) +
geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.5) +
  geom_text(aes(label = format(number_of_ride, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 1.7) + 
  
  labs(x = "Route", y = "Number of Rides", 
       title = "Cyclistic - Top 10 Most Frequent Routes", 
       subtitle =  "For Member Users") +
  scale_fill_manual(values = c("#c1c1f3","#ffd1dc")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0,10000)) +
  theme_classic(base_size = 10) + 
  theme(axis.text.y = element_text(size = 6)) + 
  coord_flip()

ggsave("Top_10_Most_Frequent_Routes_For_Member_Users.png", plot = last_plot(), dpi = 300)
```
## Top 10 Most Frequent Routes for Casual User
```{r}
final_data <- final_data %>%
  mutate(route = paste(start_station_name, "->", end_station_name))

# group by the new route column, filter for list of top 10 routes 
top_routes_member <- final_data %>%
  filter(member_casual == "casual") %>%
  group_by(route, member_casual, round_trip) %>%
  summarize(number_of_ride = n()) %>%
  ungroup() %>%
  arrange(desc(number_of_ride)) %>% 
  slice_max(order_by = number_of_ride, n = 10)

# plot the top 10 routes
ggplot(top_routes_member, aes(x = reorder(route, number_of_ride), y = number_of_ride, fill = round_trip)) +
geom_bar(stat = "identity", position = position_dodge(preserve = 'single'), width = 0.5) +
  geom_text(aes(label = format(number_of_ride, big.mark = ",")), position = position_dodge(width = 0.7), hjust = -0.1, size = 1.7) + 
  
  labs(x = "Route", y = "Number of Rides", 
       title = "Cyclistic - Top 10 Most Frequent Routes", 
       subtitle =  "For Casual Users") +
  scale_fill_manual(values = c("#c1c1f3","#ffd1dc")) +
  scale_y_continuous(labels = scales::comma_format(), limits = c(0,10000)) +
  theme_classic(base_size = 10) + 
  theme(axis.text.y = element_text(size = 6)) + 
  coord_flip()

ggsave("Top_10_Most_Frequent_Routes_For_Casual_Users.png", plot = last_plot(), dpi = 300)
```
